Modeling Galaxy Spectral Energy Distributions in the SDSS Spectroscopic Survey

### 2002-02-14 docs/paper/paper.tex
added speck
git-svn-id: svn+ssh://howdy.physics.nyu.edu/usr/local/svn/kcorrect/trunk@39 75dff457-8751-0410-ab1a-cd701654ce6b

We extract one-dimensional spectra from the two-dimension images using
a pipeline created specifically for the SDSS instrumentation
(\citealt{schlegel02a}). We also use redshifts determined by the same
pipeline; the official SDSS redshifts are obtained from a different
pipeline (\citealt{subbarao02a}). The two independent versions provide
a consistency check on the redshift determination. They are consistent
(for galaxies) at over the 99\% level.\footnote{The official pipeline
tends to work better for objects with unusual spectra, such as certain
types of stars and QSOs.}.

We use two types of magnitudes determined by the SDSS. First, the
modified form of the magnitude described by \citet{petrosian76a}, as
described in \citet{strauss02a}.  These magnitudes have a nearly
constant metric aperture as a function redshift; in addition, all the
bands use the same aperture, so the measured SED corresponds (to
within the effects of seeing) to the SED of an identifiable region of
the galaxy. However, for faint objects, the Petrosian magnitudes tend
to become noisy. Thus, for galaxies with $z>0.25$ we instead use the
higher signal-to-noise ``model magnitudes.'' Model magnitudes are
calculated in all bands using a single weighted aperture. The weighted
aperture is the better fitting model (pure exponential or pure de
Vaucouleurs) to the galaxy image in the $r$-band. In
\citet{stoughton01a} we are explicitly warned (in paragraphs written
by the lead author of this paper!)  not to mix Petrosian and model
magnitudes in a single analysis, since they measure galaxies in very
different ways. Nevertheless, for the purposes of contraining galaxy
template SEDs, it is perfectly acceptable to use {\it any}
well-defined part of any galaxy.  We reddening-correct both types of
magnitudes using the dust maps of \citet{schlegel98a}.

@@

For the purposes of applying the method described in Section
\ref{sedfit} to SDSS data, we take a subsample of the data consisting
of around 25,000 objects in the range $0.0<z<0.5$. The sample includes
both the main sample and the LRGs; we sample these spectra such that
we get an approximately even distribution within our range of
redshifts. In addition, we add results from galaxies in several plates
(totalling about 1,000 objects) which were selected by a photometric
redshift algorithm ({\bf ref who?}) to be at around $z\sim
0.3$--$0.4$. These objects are invaluable for tying down the blue end
of the templates.

@@

We choose to fit for $N_t = 4$ eigentemplates, the maximum one can use
and still fit allow freedom to fit for the templates themselves. Since
galaxies live in a low-dimensional space, we find it is worth being
able to fit for the templates. Simply using five templates (which
obviously reproduces all the magnitudes exactly) tends to yield
unphysical trends of galaxy SED versus redshift ({\it cf.}  Figure
\ref{constantz} below). Using three templates does nearly as well as
four templates in the sense that the resulting templates reproduce the
$griz$ magnitudes nearly as well. However, the fourth template is
necessary to recover the $u$-band flux to better than about 15\%. In
addition, since one of the applications of these SED determinations is
the distribution of galaxy colors in fixed-frame magnitudes, we don't
want to artificially reduce the dimensionality of the color space to
only two.

@@

An important test of the reasonableness of the fits is to check that
for a fixed type of galaxy, the distribution of the fixed frame colors
depends little on redshift. Because the galaxies shown in Figures
\ref{k_coeffdist_plot} and \ref{k_model_plot} are inhomogeneously
selected --- some are Main Sample

We must be careful in performing this check that
we are comparing Figures \ref{coeffs1}--\ref{coeffs4} demonstrates
this to be the case for galaxies in several luminosity ranges. The
left column of Figure \ref{coeffs1} shows the three coefficient ratios
$a_i/a_0$ for each galaxy as a function of redshift. The right column
shows the distribution of these coefficient ratios for two redshift
ranges, showing that they are nearly consistent. The resulting change
in the color distribution inferred from these coefficients is
similarly small, at the level of what one would expect intrinsic
evolution to cause.

In short, these fits to galaxy SEDs provide estimates which reproduce
the galaxy photometry nearly to the level of the errors in the
photometry itself, seem physically reasonable, and are consistent over
the range of redshifts we consider ($0.0. < z < 0.5$).

@@

With a model of the rest-frame galaxy SED in hand, it is 
straightforward to project that SED onto a filter corresponding to 
an observation at any redshift, and predict the so-called
$K$-correction to the galaxy magnitude in order to compare galaxies 
at different redshifts. One simply calculates the AB magnitude at 
any redshift as:
\begin{equation}
m_{AB} = -2.41 - 2.5 \log_{10}\left[
\frac{\int_{0}^{\infty} d\lambda \frac{\lambda}{1+z}
f(\frac{\lambda}{1+z}) R(\lambda)}
{\int_{0}^{\infty} d\lambda \lambda^{-1} R(\lambda)}\right].
\end{equation}
The simple way to think about this equation is that simply moving away
from the Earth doing does not change the bolometric flux of the
galaxy. So when the spectrum is stretched by the $(1+z)$ redshift
factor, as in $f(\lambda/(1+z))$, the bolometric flux should be
constant; that is, you must apply the extra multiplicative factor
$1/(1+z)$ to counteract the stretching of the spectrum. Once you have
have accounted for the fact that you are observing a particular
bandpass given the redshift of the object, you can then go ahead and
apply to the calculated magnitude the cosmological distance modulus
(that is, the luminosity distance-squared law, as tabulated by, {\it
e.g.} \citealt{hogg97a}).

>>>>

We extract one-dimensional spectra from the two-dimension images using
a pipeline created specifically for the SDSS instrumentation
(\citealt{schlegel02a}), which also fits for the redshift of each
spectrum. The official SDSS redshifts are obtained from a different
pipeline (\citealt{subbarao02a}). The two independent versions provide
a consistency check on the redshift determination. They are consistent
(for galaxies) at over the 99\% level.\footnote{The official pipeline
tends to work better for objects with unusual spectra, such as certain
types of stars and QSOs.}.

We use two types of magnitudes determined by the SDSS. First, the
modified form of the magnitude described by \citet{petrosian76a}, as
described in \citet{strauss02a}.  These magnitudes have a nearly
constant metric aperture as a function redshift; in addition, all the
bands use the same aperture, so the measured SED corresponds (to
within the effects of seeing) to the SED of an identifiable region of
the galaxy. However, for faint objects, the Petrosian magnitudes tend
to become noisy. Thus, for galaxies with $z>0.25$ we instead use the
higher signal-to-noise ``model magnitudes.'' Model magnitudes are
calculated in all bands using a single weighted aperture. The weighted
aperture is the better fitting model (pure exponential or pure de
Vaucouleurs) to the galaxy image in the $r$-band. In
\citet{stoughton01a} we are explicitly warned (in paragraphs written
by the lead author of this paper!)  not to mix Petrosian and model
magnitudes in a single analysis, since they measure galaxies in very
different ways. Nevertheless, for the purposes of contraining galaxy
template SEDs, it is perfectly acceptable to use {\it any}
well-defined part of any galaxy.  For our purposes, it is more
important to have high signal-to-noise measurement for any
well-defined part of our galaxies, even if the region of the galaxies
probed and their total estimated fluxes are inconsistent with
redshift. We extinction-correct both types of magnitudes using the
dust maps of \citet{schlegel98a}.

@@

For the purposes of applying the method described in Section
\ref{sedfit} to SDSS data, we take a subsample of the data consisting
of around 25,000 objects in the range $0.0<z<0.5$. The sample includes
both the main sample and the LRGs; we sample these spectra such that
we get an approximately even distribution within our range of
redshifts. In addition, we add results from galaxies in several plates
(totalling about 1,000 objects) which were selected by a photometric
redshift algorithm ({\bf ref who?}) to be at around $z\sim
0.3$--$0.4$. These objects are invaluable for tying down the blue end
of the templates. Finally, we exclude galaxies in the redshift range
$0.28 <z<0.32$ from the fit for the templates, for reasons which we
will explain more fully below (nevertheless, we still can and do use
the resulting templates to analyze galaxies in this range).

@@

We choose to fit for $N_t = 4$ eigentemplates, the maximum one can use
and still allow freedom to fit for the templates themselves. Since
galaxies live in a low-dimensional space, we find it is worth being
able to fit for the templates. Simply using five templates (which
obviously reproduces all the magnitudes exactly) tends to yield
unphysical trends of galaxy SED versus redshift ({\it cf.}  Figure
\ref{constantz} below). Using three templates does nearly as well as
four templates in the sense that the resulting templates reproduce the
$griz$ magnitudes nearly as well. However, the fourth template is
necessary to recover the $u$-band flux to better than about 15\%. In
addition, since one of the applications of these SED determinations is
the distribution of galaxy colors in fixed-frame magnitudes, we don't
want to artificially reduce the dimensionality of the color space to
only two.

@@

An important test of the reasonableness of the fits is to check that
for a fixed type of galaxy, the distribution of the fixed frame colors
depends little on redshift. Because the galaxies shown in Figures
\ref{k_coeffdist_plot} and \ref{k_model_plot} are inhomogeneously
selected --- some are Main Sample galaxies, some are Luminous Red
Galaxies, and some are selected from the photometric redshift plates
--- we will split our sample into two well-defined sets of
objects. First, we choose a set of Main Sample galaxies in the
luminosity range $-21.5 < M_{\band{0.1}{r}} < -21$. We reconstruct the
four colors $^{0.1}(u-g)$, $^{0.1}(g-r)$, $^{0.1}(r-i)$, and
$^{0.1}(i-z)$ for these galaxies and plot the reconstructions as a
function of redshift in left-hand panels of Figure
\ref{main_colors_plot}. The plots show a general consistency in the
fixed-frame color distribution of these objects with redshift. The
right-hand panels show the distribution for redshifts $z<0.13$ (solid
histogram) and $z>0.13$ (dotted histogram) . Small changes are
discernible in the distributions, mostly attributable to the increased
errors at higher redshift. Note that for $z<0.1$ the $^{0.1}(u-g)$
color depends on an extrapolation of the SED in the blue, while for
$z>0.1$ the $^{0.1}(i-z)$ color depends on an extrapolation of the SED
in the red.

Second, we choose a set of Cut I LRGs with $-22.8 < M_{\band{0.1}{r}}
< -22.5$. Now we reconstruct the colors to a fixed-frame corresponding
to the rest-frame bands used to observe a galaxy at $z=0.3$.  Near
$z=0.3$ there is some singular behavior of the colors, which we will
discuss in more detail later. Again, for $z<0.3$ the $^{0.3}(u-g)$
color relies on an extrapolation, which is probably responsible for
the scatter in that color at low redshifts. Otherwise, the
distribution of colors is remarkably constant. $^{0.3}(g-r)$ (which is
very similar to $^{0.0}(u-g)$) experiences a blueward shift of about
0.1 magnitudes between $z=0.1$ and $z=0.45$, which is attributable to
passive evolution.

In short, these fits to galaxy SEDs provide estimates which reproduce
the galaxy photometry nearly to the level of the errors in the
photometry itself, seem physically reasonable, and are consistent over
the range of redshifts we consider ($0.0 < z < 0.5$). The only
remaining problem is the aforementioned difficulties at $z=0.3$,
discussed in more detail in the next section.

\section{The $g$/$r$ Gap}
\label{grgap}

It is clear from Figure \ref{k_kcorrect_plot} that $z=0.3$ is a
special redshift regime where our fits become unphysical. In fact, it
is easy to demonstrate that the diagonal spur (going from the upper
left to the lower right) around the red dot in the upper right panel
of Figure \ref{k_coeffdist_plot} is due to galaxies at $z\sim
0.3$. The difficulty arises because the 4000 \AA\ break lies in the
gap between $g$ and $r$ at $z=0.3$. The algorithm requires strong
gradients in the spectrum in this area to explain the 4000\AA break
--- however, the gap in the bands causes a degeneracy arise.

To demonstrate this degeneracy, Figure \ref{spur} shows the spectrum
corresponding to the direction of the spur in the $a_1/a_0$-$a_3/a_0$
plane of Figure \ref{k_coeffdist_plot}. We overplot the \band{0.3}{g}
and \band{0.3}{r} bands. The spur has a strong peak in the gap between
the two bands at $z=0.3$. Because of this gap, at $z\sim 0.3$ the
algorithm tends to use this direction to help fit the $u$, $i$, and
$z$ fluxes, with relatively little harm done to the $g$ and $r$
fits. This explains why the reconstructed fluxes in Figure
\ref{k_model_plot} fit well even at $z=0.3$, even though the spur at
$z=0.3$ in the distribution of coefficients is so evident.  This
difficulty is why, in Section \ref{data}, we exclude the regime around
$z\sim 0.3$; otherwise the templates became distracted by the
degeneracies in this regime.

The question arises of what to do about this problem. One solution is
to simply work around it. It is clear we cannot use the fits presented
here to $K$-correct galaxies which are at $z=0.3$ to other redshifts
reliably. However, nothing prevents one from $K$-correcting galaxies
at other redshifts {\it to} $z=0.3$. If you are considering the LRG
sample, the median redshift is close to $z=0.3$ anyway, so this is a
viable solution (and the one adopted in Figure \ref{lrg_colors_plot}).

A second solution, not adopted here, is to place stronger constraints
in the SED fit on the form of the 4000 \AA\ break. One obvious way of
doing this is to use the spectra to constrain the size of the break
(making it a weak constraint to prevent biases). This constraint could
be applied in a linear fashion to retain the speed of our method ---
except for the fact that you would have to read in the spectra to
apply the $K$-correction. For simplicity, and because the aforemention
work-around is so easy to obey, we have not implemented this solution
here.

@@

With a reconstruction of the rest-frame galaxy SED in hand, it is
straightforward to project that SED onto a filter corresponding to an
observation at any redshift, and predict the so-called $K$-correction
to the galaxy magnitude in order to compare galaxies at different
redshifts. One simply calculates the AB magnitude at any redshift as:
\begin{equation}
m_{AB,\mathrm{rec}} = -2.41 - 2.5 \log_{10}\left[
\frac{\int_{0}^{\infty} d\lambda \frac{\lambda}{1+z}
f_{\mathrm{rec}}(\frac{\lambda}{1+z}) R(\lambda)}
{\int_{0}^{\infty} d\lambda \lambda^{-1} R(\lambda)}\right].
\end{equation}
The simple way to think about this equation is that simply moving away
from the Earth doing does not change the bolometric flux of the
galaxy. So when the spectrum is stretched by the $(1+z)$ redshift
factor, as in $f(\lambda/(1+z))$, the bolometric flux should be
constant; that is, you must apply the extra multiplicative factor
$1/(1+z)$ to counteract the stretching of the spectrum. Once you have
have accounted for the fact that you are observing a particular
bandpass given the redshift of the object, you can then go ahead and
apply to the calculated magnitude the cosmological distance modulus
(that is, the luminosity distance-squared law, as tabulated by, {\it
e.g.} \citealt{hogg97a}).
